The Radio Sky at Meter Wavelengths: $m$-Mode Analysis Imaging\\ with the Owens Valley Long Wavelength Array

### 2017-09-21 paper.tex
this is the draft I'm taking to gregg's house

At redshifts $20 \gtrsim z \gtrsim 7$ the 21~cm hyperfine structure line of neutral hydrogen is
expected to produce a 10 to 100 mK perturbation in the CMB spectrum \citep{2006PhR...433..181F,
2012RPPh...75h6901P}. The amplitude of this perturbation on a given line-of-sight is a function of
the neutral fraction of hydrogen, the baryon overdensity, the spin temperature relative to the CMB
temperature at the given redshift, and the line-of-sight peculiar velocity of the gas.  The spatial
power spectrum of this perturbation is thought to be dominated by inhomogeneous heating of the IGM
at $z\sim 20$ \citep{2014MNRAS.437L..36F}, and by growing ionized bubbles during the EoR at $z\sim
7$ where a detection can constrain the ionizing efficiency of early galaxies, the UV photon
mean-free-path, and the minimum halo mass that can support star formation
\citep{2015MNRAS.449.4246G}.

@@

It can be shown that there is a discrete linear relationship between the measured $m$-modes
$V^{ij}_{m,\nu}$ and the spherical harmonic coefficients of the sky brightness $a_{lm,\nu}$.
\begin{equation}\label{eq:m-mode-sum-equation}
    V^{ij}_{m,\nu} = \sum_l B^{ij}_{lm,\nu} a_{lm,\nu}\,,
\end{equation}
where the transfer coefficients $B^{ij}_{lm,\nu}$ define the interferometer's response to the sky.
These transfer coefficients are computed from the spherical harmonic transform of the baseline
transfer function defined by Equation~\ref{eq:baseline-transfer-function}.

@@

The Owens Valley Long Wavelength Array (OVRO-LWA) is a 288-element interferometer located at the
Owens Valley Radio Observatory near Big Pine, California \citep{todo_hallinan_2017}.  The OVRO-LWA
is a low-frequency instrument with instantaneous bandwidth covering 27.384 MHz to 84.912 MHz with 24
kHz channelization.  Each antenna stand hosts two perpendicular broadband dipoles so that there are
$288\times2$ signal paths in total. These signal paths feed into the 512-input LEDA correlator
\citep{2015JAI.....450003K}, which allows the OVRO-LWA to capture the entire visible hemisphere in a
single snapshot image.  In the current configuration 32 antennas (64 signal paths) are unused.

The 288 antennas are arranged in a pseudo-random configuration optimized to minimize sidelobes in
snapshot imaging.  251 of the antennas are contained within a 200 m diameter core. 32 antennas are
placed outside of the core in order to extend the maximum baseline length to $\sim$1.5~km. The final
5 antennas are equipped with noise-switched front ends for calibrated total power measurements of
the global sky brightness.  These antennas are used as part of the LEDA experiment
\citep{todo_price_2017} to measure the global signal of 21~cm absorption from the cosmic dawn.
Figure~\ref{fig:antenna-layout} is a diagram of the antenna configuration.

@@

\subsection{Source Removal}

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/spectral-index/internal-spectral-index}
    \caption{
        This Mollweide-projected map is constructed from 3 maps of the sky at 36.528~MHz (red),
        52.224~MHz (green), and 73.152~MHz (blue). The maps are scaled by $\nu^{2.5}$ before
        combining and the color scale is logarithmic (as in Figure~\ref{fig:channel-maps}). The
        local spectral index (estimated by means of a local T-T plot) is overlaid as a contour plot.
        \todo{
            The spectral index near the galactic plane is very wrong because the local T-T plot is
            contaminated with multiple types of emission. I need to redo this plot by replacing the
            galactic plane with hand-picked regions for estimating the spectral index.
        }
    }
    \label{fig:internal-spectral-index}
\end{figure*}

@@

Eight sky maps were constructed using Tikhonov-regularized $m$-mode analysis imaging and cleaning
with observations from the OVRO-LWA. These maps are shown in Figure~\ref{fig:channel-maps}. Each map
covers the sky north of $\delta=-30^\circ$ with $\sim 15$ arcmin angular resolution resolution. The
8 brightest northern hemisphere point sources are removed from each map (Cyg~A, Cas~A, Vir~A, Tau~A,
Her~A, Hya~A, Per~B, and 3C~353). There is a small blank region near $l=+45.7^\circ$,
$b=-47.9^\circ$ corresponding to the position of the Sun during the observing window. The properties
of each map, including frequency, bandwidth, angular resolution, and thermal noise are presented in
Table~\ref{tab:summary}.

@@

In order to suppress sources of terrestrial interference, all spherical harmonics with $m=0$, or
$m=1$ and $l>100$ are filtered from the map (where the spherical harmonics are defined in the J2017
coordinate system). As will be discussed in \S\ref{sec:rfi}, these spherical harmonics are
particularly susceptible to contamination by RFI and common-mode problems. As a consequence,
astronomical emission that circles the J2017 north celestial pole (NCP) is filtered from the maps.
This tends to create rings around the NCP at the declination of bright point sources. These rings
are naturally removed from the map during CLEANing as long as this filtering step is including in
the PSF calculation.

@@

One of the key assumptions made by $m$-mode analysis is that the sky is completely static.  We
assume that the only time-dependent behavior is the rotation of the Earth, which slowly rotates the
sky through the fringe patterns of the interferometer. At low frequencies the ionosphere violates
this assumption. In particular, ionospheric scintillation will cause even static sources to exhibit
significant variability (see Figure~\ref{fig:scintillation}).

@@

\begin{figure*}[t]
    \centering
    \includegraphics[height=0.32\textheight]{figures/haslam/haslam-spectral-index}
    \caption{
        The spectral index between the reprocessed Haslam 408 MHz map and the 73.152 MHz map
        (degraded to 56 arcmin resolution). The white regions of the plot correspond to areas where
        there was a negative correlation between the brightness temperature at 73.152 MHz and at 408
        MHz. This can either be due to the presence of multiple emission mechanisms in close
        proximity to each other or due to insufficient dynamic range in temperature leading to
        effectively no correlation between the two temperatures. Regardless, a single power law is
        insufficient to describe the data in these areas.
        \todo{
            Cas A is the extremely blue feature in this plot. It needs to be blotted out because
            it's not present in my maps, but it is present in the Haslam maps. For whatever reason
            this isn't a problem with Cyg A.
        }
    }
    \label{fig:haslam-spectral-index}
\end{figure*}

The Haslam 408 MHz map \citep{1981A&A...100..209H, 1982A&AS...47....1H} was reprocessed by
\citet{2015MNRAS.451.4311R} to remove artifacts associated with $1/f$ noise and bright sources. In
order to compare the OVRO-LWA maps with the lower resolution Haslam map, the OVRO-LWA maps were
degraded to 56 arcmin resolution and the spectral index between the 73.152 MHz map and the 408 MHz
Haslam map was computed by means of a T-T plot at each pixel in the maps (in order to obviate the
need for a zero-level correction to the 73.152 MHz map). The results of this computation are shown
in Figure~\ref{fig:haslam-spectral-index}. Because this comparison uses a local T-T plot to estimate
the spectral index, the angular resolution of this computation is somewhat degraded relative to
other comparisons in the literature but the spectral indices in
Figure~\ref{fig:haslam-spectral-index} are robust to errors in the zero-level in either map.

@@

At frequencies above $\sim55$~MHz, the angular resolution of these maps is limited by the selection
of $l_\text{max}=1000$. Future work will increase $l_\text{max}$ to remove this restriction, as well
as include more time and bandwidth to improve the thermal noise. The usage of nighttime only data
can help mitigate dynamic range limitations from the ionosphere and also prevent the Sun from
showing up in the maps. Observations could also be extended to slightly higher and lower frequencies
($\sim27$ to $85$~MHz) to take advantage of the full frequency range of the OVRO-LWA. The higher
frequencies are particularly interesting in order to maximize the overlap with the MWA in the
southern hemisphere, which could be used to fill-in the hole around the southern celestial pole.

>>>>

At redshifts $20 \gtrsim z \gtrsim 7$ the 21~cm hyperfine structure line of neutral hydrogen is
expected to produce a 10 to 100 mK perturbation in the Cosmic Microwave Background (CMB) spectrum
\citep{2006PhR...433..181F, 2012RPPh...75h6901P}. The amplitude of this perturbation on a given
line-of-sight is a function of the neutral fraction of hydrogen, the baryon overdensity, the spin
temperature relative to the CMB temperature at the given redshift, and the line-of-sight peculiar
velocity of the gas.  The spatial power spectrum of this perturbation is thought to be dominated by
inhomogeneous heating of the IGM at $z\sim 20$ \citep{2014MNRAS.437L..36F}, and by growing ionized
bubbles during the EoR at $z\sim 7$ where a detection can constrain the ionizing efficiency of early
galaxies, the UV photon mean-free-path, and the minimum halo mass that can support star formation
\citep{2015MNRAS.449.4246G}.

@@

\citet{2014ApJ...781...57S, 2015PhRvD..91h3514S} show that there is a discrete linear relationship
between the measured $m$-modes $V^{ij}_{m,\nu}$ and the spherical harmonic coefficients of the sky
brightness $a_{lm,\nu}$.
\begin{equation}\label{eq:m-mode-sum-equation}
    V^{ij}_{m,\nu} = \sum_l B^{ij}_{lm,\nu} a_{lm,\nu}\,,
\end{equation}
where the transfer coefficients $B^{ij}_{lm,\nu}$ define the interferometer's response to the sky.
These transfer coefficients are computed from the spherical harmonic transform of the baseline
transfer function defined by Equation~\ref{eq:baseline-transfer-function}.

@@

The Owens Valley Long Wavelength Array (OVRO-LWA) is a 288-element interferometer located at the
Owens Valley Radio Observatory near Big Pine, California \citep{todo_hallinan_2017}.  The OVRO-LWA
is a low-frequency instrument with instantaneous bandwidth covering 27.384 MHz to 84.912 MHz and
with 24 kHz channelization.  Each antenna stand hosts two perpendicular broadband dipoles so that
there are $288\times2$ signal paths in total. These signal paths feed into the 512-input LEDA
correlator \citep{2015JAI.....450003K}, which allows the OVRO-LWA to capture the entire visible
hemisphere in a single snapshot image.  In the current configuration 32 antennas (64 signal paths)
are unused.

The 288 antennas are arranged in a pseudo-random configuration optimized to minimize sidelobes in
snapshot imaging (see Figure~\ref{fig:antenna-layout}).  251 of the antennas are contained within a
200 m diameter core. 32 antennas are placed outside of the core in order to extend the maximum
baseline length to $\sim$1.5~km. The final 5 antennas are equipped with noise-switched front ends
for calibrated total power measurements of the global sky brightness.  These antennas are used as
part of the LEDA experiment \citep{todo_price_2017} to measure the global signal of 21~cm absorption
from the cosmic dawn.

@@

\subsection{Source Removal}\label{sec:source-removal}

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/spectral-index/internal-spectral-index}
    \caption{
        This Mollweide-projected map is constructed from 3 maps of the sky at 36.528~MHz (red),
        52.224~MHz (green), and 73.152~MHz (blue). The maps are scaled by $\nu^{2.5}$ before
        combining and the color scale is logarithmic (as in Figure~\ref{fig:channel-maps}). The
        local spectral index (estimated by means of a local T-T plot) is overlaid as a contour plot.
    }
    \label{fig:internal-spectral-index}
\end{figure*}

@@

Eight sky maps were constructed using Tikhonov-regularized $m$-mode analysis imaging and cleaning
with observations from the OVRO-LWA. These maps are shown in Figure~\ref{fig:channel-maps}. Each map
covers the sky north of $\delta=-30^\circ$ with $\sim 15$ arcmin angular resolution resolution. The
8 brightest northern hemisphere point sources are removed from each map (Cyg~A, Cas~A, Vir~A, Tau~A,
Her~A, Hya~A, Per~B, and 3C~353) as describe in \S\ref{sec:source-removal}. There is a small blank
region near $l=+45.7^\circ$, $b=-47.9^\circ$ corresponding to the position of the Sun during the
observing window. The properties of each map, including frequency, bandwidth, angular resolution,
and thermal noise are presented in Table~\ref{tab:summary}.

@@

In order to suppress sources of terrestrial interference, all spherical harmonics with $m=0$, or
$m=1$ and $l>100$ are filtered from the map (where the spherical harmonics are defined in the J2017
coordinate system). As will be discussed in \S\ref{sec:rfi}, these spherical harmonics are
particularly susceptible to contamination by RFI and common-mode pickup. As a consequence,
astronomical emission that circles the J2017 north celestial pole (NCP) is filtered from the maps.
This tends to create rings around the NCP at the declination of bright point sources. These rings
are naturally removed from the map during CLEANing as long as this filtering step is including in
the PSF calculation.

@@

One of the key assumptions made by $m$-mode analysis is that the sky is completely static.  We
assume that the only time-dependent behavior is the rotation of the Earth, which slowly rotates the
sky through the fringe patterns of the interferometer. At low frequencies the ionosphere violates
this assumption. In particular, ionospheric scintillation and refractive offsets will cause even
static sources to exhibit significant variability (see Figure~\ref{fig:scintillation}).

@@

\begin{figure*}[t]
    \centering
    \includegraphics[height=0.32\textheight]{figures/haslam/haslam-spectral-index}
    \caption{
        The contours show the spectral index between the reprocessed Haslam 408 MHz map and the
        73.152 MHz map (degraded to 56 arcmin resolution). These contours are overlaid on the same
        3-color map as in Figure~\ref{fig:internal-spectral-index}. The regions with shallow
        spectral index between 73.152 MHz and 408 MHz coincide with regions with large free-free
        absorption that is evident between 36.528 MHz and 73.152 MHz.
    }
    \label{fig:haslam-spectral-index}
\end{figure*}

The Haslam 408 MHz map \citep{1981A&A...100..209H, 1982A&AS...47....1H} was reprocessed by
\citet{2015MNRAS.451.4311R} to remove artifacts associated with $1/f$ noise and bright sources. In
order to compare the OVRO-LWA maps with the lower resolution Haslam map, the OVRO-LWA maps were
degraded to 56 arcmin resolution and the spectral index between the 73.152 MHz map and the 408 MHz
Haslam map was computed by means of a local T-T plot at each pixel in the maps (in order to obviate
the need for a zero-level correction to the 73.152 MHz map). The results of this computation are
shown in Figure~\ref{fig:haslam-spectral-index}. Because this comparison uses a local T-T plot to
estimate the spectral index, the angular resolution of this computation is somewhat degraded
relative to other comparisons in the literature but the spectral indices in
Figure~\ref{fig:haslam-spectral-index} are robust to zero-level errors in either map.

@@

At frequencies above $\sim55$~MHz, the angular resolution of these maps is limited by the selection
of $l_\text{max}=1000$. Future work will increase $l_\text{max}$ to remove this restriction, as well
as include more time and bandwidth to improve the thermal noise. The usage of nighttime only data
can help mitigate dynamic range limitations from the ionosphere and also eliminate solar sidelobe
residuals. Observations could also be extended to slightly higher and lower frequencies ($\sim27$ to
$85$~MHz) to take advantage of the full frequency range of the OVRO-LWA. The higher frequencies are
particularly interesting in order to maximize the overlap with the MWA in the southern hemisphere,
which could be used to fill-in the hole around the southern celestial pole.
